---
title: "Comparison of Air pollution exposure models - RAQDPS-FW and CanOSSEM"
output: html_document
---

## Dependencies/Libraries

```{r setup, message = FALSE, warning=FALSE, echo=FALSE}
library(tidyverse)
library(lubridate)
library(stringr)
library(scales) 
library(ggthemes)
library(heemod)
require(sf)

print("hi")
```

# Importing data

```{r}
canossem <- read.csv("./data/canossem_mo_hsda_2012_2022.csv")
raqdps <- read.csv("./data/raqdps_mo_hsda_2012_2023.csv")
raqdpsfw <- read.csv("./data/raqdpsfw_mo_hsda_2012_2023.csv")


glimpse(canossem)
```

# Data cleaning and merging models

```{r addingSpatialData, message = FALSE, warning=TRUE, echo=FALSE}
# Now merging with map:
canossem <- canossem %>% 
  mutate(HLTH_SERVICE_DLVR_AREA_CODE = as.character(HLTH_SERVICE_DLVR_AREA_CODE)) %>%
  select(c("HLTH_SERVICE_DLVR_AREA_CODE", "year", "month", "monthly_smoke", "date")) %>%
  mutate(date = as.Date(date, format = "%Y-%m-%d")) %>%
  filter(date >= as.Date("2018-01-01") & date <= as.Date("2022-12-31")) %>%
  rename("average_mo" = "monthly_smoke") %>%
  left_join(bcmaps::health_hsda(), by="HLTH_SERVICE_DLVR_AREA_CODE")
glimpse(canossem)

raqdps <- raqdps %>%
  select(c("HLTH_SERVICE_DLVR_AREA_NAME", "year", "month", "average_mo", "date")) %>%
  mutate(date = as.Date(date, format = "%Y-%m-%d")) %>%
  filter(date >= as.Date("2018-01-01") & date <= as.Date("2022-12-31")) %>%
  left_join(bcmaps::health_hsda(), by="HLTH_SERVICE_DLVR_AREA_NAME")

glimpse(raqdps)

raqdpsfw <- raqdpsfw %>%
    select(c("HLTH_SERVICE_DLVR_AREA_NAME", "year", "month", "average_mo", "date")) %>%
  mutate(date = as.Date(date, format = "%Y-%m-%d")) %>%
  filter(date >= as.Date("2018-01-01") & date <= as.Date("2022-12-31")) %>%
  left_join(bcmaps::health_hsda(), by="HLTH_SERVICE_DLVR_AREA_NAME")

glimpse(raqdpsfw)

# Combining

canossem <- canossem %>%
  mutate(Dataset = "canossem")

raqdpsfw <- raqdpsfw %>%
  mutate(Dataset = "raqdpsfw")

combined_can_raq <- bind_rows(canossem, raqdpsfw)

joined_can_raq <- canossem %>%
  select(c("average_mo", "date", "HLTH_SERVICE_DLVR_AREA_NAME")) %>%
  inner_join(raqdpsfw, by = c("date", "HLTH_SERVICE_DLVR_AREA_NAME"), suffix = c("_canossem", "_raqdpsfw"))

glimpse(joined_can_raq)
```

# Plot of average monthly PM2.5 values by HSDA (CanOSSEM vs. RAQDPS-FW)

```{r, echo=FALSE, warning=FALSE, message=FALSE}
#| label: fig-smoke
#| fig-cap: Daily Smoke Exposure Levels (PM2.5) across BC



haLineMo <- function (data, title) {
    ggplot(data) +
    geom_line(aes(x = date, y = average_mo, colour = HLTH_SERVICE_DLVR_AREA_NAME)) +
    scale_x_date(date_labels = "%m-%Y", date_breaks = "6 month") +
    theme_few(base_size = 8) +
    theme(
      axis.text.x = element_text(angle = 90, hjust = 1),
      legend.position = "none"
    ) +
    facet_wrap(~ HLTH_SERVICE_DLVR_AREA_NAME) +
    labs(title = title,
    x = "",
    y = "µg/m³")
}


haLineMo(canossem, "CanOSSEM Monthly")
haLineMo(raqdps, "RAQDPS Monthly")
haLineMo(raqdpsfw, "RAQDPS-FW Monthly")



haLineMoCombined <- function(data, title) {
  ggplot(data) +
    geom_line(aes(x = date, y = average_mo, colour = Dataset), linewidth = 0.5, alpha = 0.7) +
    scale_x_date(date_labels = "%m-%Y", date_breaks = "6 month") +
    scale_colour_manual(values = c(raqdpsfw = "blue3", canossem= "darkorange")) +
    theme_few(base_size = 8) +
    theme(
      axis.text.x = element_text(angle = 90, hjust = 1),
      legend.position = "right"
    ) +
    facet_wrap(~ HLTH_SERVICE_DLVR_AREA_NAME) +
    labs(title = title,
    x = "",
    y = "µg/m³")
}

plot1 <- haLineMoCombined(combined_can_raq, "CanOSSEM vs. RAQDPS-FW")

ggsave("CanOSSEM vs. RAQDPS-FW, 2018-2022.png", plot1, width = 6, height = 4)

haLineMoCombinedLog <- function(data, title) {
  ggplot(data) +
    geom_line(aes(x = date, y = log(average_mo), colour = Dataset), linewidth = 0.5) +
    scale_x_date(date_labels = "%m-%Y", date_breaks = "6 month") +
    theme_few(base_size = 8) +
    theme(
      axis.text.x = element_text(angle = 90, hjust = 1),
      legend.position = "right"
    ) +
    facet_wrap(~ HLTH_SERVICE_DLVR_AREA_NAME) +
    labs(title = title,
    x = "",
    y = "µg/m³")
}

haLineMoCombinedLog(combined_can_raq, "Log(CanOSSEM vs. RAQDPS-FW)")


```

# Summary Statistics

```{r, warning=FALSE, message=FALSE, echo=FALSE}
summary_stats1 <- combined_can_raq %>%
  group_by(Dataset, HLTH_SERVICE_DLVR_AREA_NAME) %>%
  summarise(
    mean_pm25 = mean(average_mo, na.rm = TRUE),
    median_pm25 = median(average_mo, na.rm = TRUE),
    max_pm25 = max(average_mo, na.rm = TRUE),
    min_pm25 = min(average_mo, na.rm = TRUE)
    )


summary_stats2 <- joined_can_raq %>%
  select(date, HLTH_SERVICE_DLVR_AREA_NAME, average_mo_canossem, average_mo_raqdpsfw) %>%
  mutate(pm25_diff = average_mo_canossem - average_mo_raqdpsfw)

glimpse(summary_stats2)


# Plots line plot of differences
plot2 <- ggplot(summary_stats2, aes(x = date, y = pm25_diff)) +
  geom_line(colour = "black", linewidth = 0.5) +
  geom_hline(yintercept = 0, color = "blue", linetype = "solid", linewidth = 0.3, alpha = 0.65) +
  scale_x_date(date_labels = "%m-%Y", date_breaks = "6 month") +
  theme_few(base_size = 8) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    legend.position = "right"
  ) +
  facet_wrap(~ HLTH_SERVICE_DLVR_AREA_NAME) +
  xlab("Date") +
  ylab("Difference in Monthly Smoke Exposure (µg/m³)") +
  ggtitle("Difference in Monthly Smoke Exposure Between Canossem and Raqdpsfw")

ggsave("CanOSSEM vs. RAQDPS-FW differences, 2018-2022.png", plot2, width = 6, height = 4)

# Scatter plot of difference between pm25 level

scatter_all <- ggplot(summary_stats2, aes(x = average_mo_canossem, y = average_mo_raqdpsfw)) +
  geom_point(size = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +
  theme_few() +
  xlab("CanOSSEM Monthly Smoke (ug/m³)") +
  ylab("RAQDPS-FW Monthly Smoek (ug/m³)") +
  ggtitle("Scatter Plot of Monthly Smoke Exposure Comparison - CanOSSEM vs. RAQDPS-FW")

scatter_all

scatter_hsda <- ggplot(summary_stats2, aes(x = average_mo_canossem, y = average_mo_raqdpsfw)) +
  geom_point(aes(color = HLTH_SERVICE_DLVR_AREA_NAME), size = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +
  theme_classic() +
  xlab("CanOSSEM Monthly Smoke (ug/m³)") +
  ylab("RAQDPS-FW Monthly Smoek (ug/m³)") +
  ggtitle("Scatter Plot of Monthly Smoke Exposure Comparison by HSDA - CanOSSEM vs. RAQDPS-FW")

scatter_hsda

```

# Pearson and Spearman results

## Pearson (r)

A correlation of r = 0.722 (strong) relationship between CanOSSEM and RAQDPS-FW on a monthly time frame.

The air pollution predictions are moving together linearly. Magnitude is consistent.

## Spearman (rho)

A correlation of rho = 0.010 (very weak) monotonic relationship between CanOSSEM and RAQDPS-FW modeled levels of exposure. Since spearman is a ranking correlation, the relative magnitude of PM25 estimates are not strongly correlated.

They question is whether these correlations are similar across different comparisons in literature?

```{r}
airpoll_pearson <- cor.test(joined_can_raq$average_mo_canossem, joined_can_raq$average_mo_raqdpsfw,
                             method = "pearson")

airpoll_spearman <- cor.test(joined_can_raq$average_mo_canossem, joined_can_raq$average_mo_raqdpsfw,
                             method = "spearman")

airpoll_pearson
airpoll_spearman

```

# Testing assumptions

The qqplot (testing normal distribution assumption) looks skewed to the right for both CanOSSEM and RAQDPS-FW.

A log-transform looks more appropriate for both CanOSSEM and RAQDPS-FW

```{r}
# Linearity (qqplot)
qqnorm(joined_can_raq$average_mo_canossem)

qqnorm(log(joined_can_raq$average_mo_canossem))

qqnorm(joined_can_raq$average_mo_raqdpsfw)

qqnorm(log(joined_can_raq$average_mo_raqdpsfw))
```

# Trying some regression

```{r}
lm_air <- lm(joined_can_raq$average_mo_canossem ~ joined_can_raq$average_mo_raqdpsfw)

summary(lm_air)

qqnorm(residuals(lm_air))
qqline(residuals(lm_air))


lm_air_log <- lm(log(joined_can_raq$average_mo_canossem) ~ log(joined_can_raq$average_mo_raqdpsfw))

summary(lm_air_log)
qqnorm(residuals(lm_air_log))
qqline(residuals(lm_air_log))

```

# Bland-Altman plot (difference in mean values)

```{r}
joined_can_raq2 <- joined_can_raq %>%
  mutate(mean_values = (average_mo_canossem + average_mo_raqdpsfw) / 2,
         diff_values = average_mo_canossem - average_mo_raqdpsfw)

mean_diff <- mean(joined_can_raq2$diff_values, na.rm = TRUE)
sd_diff <- sd(joined_can_raq2$diff_values, na.rm = TRUE)

upper_limit <- mean_diff + (1.96 * sd_diff)
lower_limit <- mean_diff - (1.96 * sd_diff)

ggplot(joined_can_raq2, aes(x = mean_values, y = diff_values)) +
  geom_point(alpha = 0.6) +  # Scatter plot of differences vs means
  geom_hline(yintercept = mean_diff, linetype = "solid", color = "blue") +  # Bias line
  geom_hline(yintercept = upper_limit, linetype = "dashed", color = "red") +  # Upper limit of agreement
  geom_hline(yintercept = lower_limit, linetype = "dashed", color = "red") +  # Lower limit of agreement
  labs(title = "Bland-Altman Plot: CanOSSEM vs RAQDPS-FW",
       x = "Mean of Canossem and RAQDPS-FW",
       y = "Difference between Canossem and RAQDPS-FW") +
  theme_few()
```

For the majority of data, both CanOSSEM and RAQDPS-FW seem to be in reasonable agreement, most points fall within the limits of agreement (1 sd in +/-) and produce similar results. Differences are greatest at higher pollution levels.
